---
title: "Differential Expression"
output: html_document
---
 
Use this notebook to perform differential expression and all other post-DE
analyses. 

### Setting up the correct working directory
```{r}
setwd("/projectnb/bf528/students/jawa/project-1-template-jawa23bio")
```

### Loading the necessary libraries
```{r}
library(tidyverse)
library(DESeq2)
```

### Load the counts data
```{r} 
counts_matrix <- read.csv("results/full_filtered_counts_matrix.csv", row.names = 1)

# Displaying the first 5 entries from the file
head(counts_matrix)
```

### Sample information
```{r}
coldata <- data.frame(
  sample = colnames(counts_matrix)
)
coldata$timepoint <- substr(coldata$sample, 1, 2)

coldata
```

### Create DESeqDataSet object
```{r}
dds <- DESeqDataSetFromMatrix(countData = counts_matrix, colData = coldata, design = ~ timepoint)
```

### Perform differential expression analysis
```{r}
dds <- DESeq(dds)
```

### Contrast for differential expression analysis (AD vs p0)
```{r}
contrast <- c("timepoint", "AD", "P0")
```

### DE Results for the specified contrast
```{r}
results_AD_P0 <- results(dds, contrast = contrast)
results_AD_P0
```

### DE Results in a tibble
```{r}
results <- as_tibble(results_AD_P0)
results$gene_id <- rownames(results_AD_P0)
results <- results %>% select(gene_id, everything())

results
```

## _Gene Map Information_

### Reading gene map data
```{r}
gene_map <- read.csv("results/full_genemap.csv", header = TRUE)

# Displaying the first 5 entries from the file
head(gene_map)
```

### Joining gene names to results tibble
```{r}
results <- left_join(results, gene_map, by = "gene_id")
results <- results %>% select(gene_id, gene_name, everything())

results
```